Memes: an R interface to the MEME Suite

### 2021-03-26 manuscript.Rmd
remove stuff about df

Here we present `memes`, a motif analysis package that provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor
data structures provide a full-featured representation of MEME Suite output
data, providing users quick access to all relevant data structures with simple
syntax. **TODO: Depending on length restrictions, end here**

**TODO: just end above since universalmotif_df exists now**

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow users to
integrate external data with motif databases to aid analysis, such as filtering
a set of transcription factor motifs based on gene expression in the tissue of
interest, then using that filtered set to search for motifs associated with
expressed factors. By maintaining compatability with `universalmotif`, `memes`
outputs can be readily used in a diverse set of tools beyond the `memes`
framework.

`memes` also provides simple utilities for data visualization where appropriate.
Providing R representations of MEME Suite data objects enables generation of
novel data visualizations using the well-established tools available in the R
landscape. Therefore, `memes` empowers users to deeply analyze and visualize
motif analysis data.

@@

The `universalmotif_df` data structure is a powerful R/Bioconductor
representation for motif matrices and their associated metadata
\cite{universalmotif}. `universalmotif_df` format is an alternative
representation of `universalmotif` objects where motifs are stored along rows of
a data.frame, and columns store metadata associated with each motif. Adding
one-to-many relationships within this structure is trivial, as additional
`unviersalmotif_df`s can be nested within eachother. These `universalmotif_df`s
form the basis for a majority of `memes` data outputs. These
structures are also valid input types to `memes` functions, and when used as
such, output data are appended as new columns to the input data, ensuring data
provenance.

**Figure?: showing how the data.frame connects to universalmotif**

## TODO:
 - `remove_duplicate_motifs`
 -

The universalmotif data.frames generated by memes are not limited to storing
motifs from MEME Suite data. They can also be used to manipulate or store other
descriptive properties of motifs from multiple sources. `universalmotif`
provides several import functions which can convert motifs from external
databases queried from the `MotifDb` package, or from files in a variety of
non-meme file formats \cite{motifdb, universalmotif}. Once imported in `universalmotif` format
the `memes` function `as_universalmotif_dataframe()` converts any
`universalmotif` object into a universalmotif data.frame. Similar to MEME Suite
data stored in this format, if changes are made to the linked metadata columns,
the linked universalmotif object can be updated by calling `update_motifs()` to
retain the data.frame structure for further analysis, or by calling
`as_universalmotif()` to convert the data.frame back into universalmotif format,
discarding columns that cannot be stored within `universalmotif` format.
Therefore, this data structure empowers users to import and analyze motif lists
from a diverse set of sources for use with `memes` or for simple manipulation of
motif metadata using common R tools. In this way, memes enables large-scale
motif database manipulation & analysis within the preexisting R framework.

>>>>

Here we present `memes`, a motif analysis package that provides a seamless
interface to MEME Suite tools within R. `memes` uses base R and Bioconductor
data types for data input and output, facilitating better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines without additional data processing steps. Additionally, R/Bioconductor
data structures provide a full-featured representation of MEME Suite output
data, providing users quick access to all relevant data structures with simple
syntax.

@@

The `universalmotif_df` data structure is a powerful R/Bioconductor
representation for motif matrices and their associated metadata
\cite{universalmotif}. `universalmotif_df` format is an alternative
representation of `universalmotif` objects where motifs are stored along rows of
a data.frame, and columns store metadata associated with each motif. Adding
one-to-many relationships within this structure is trivial, as additional
`unviersalmotif_df`s can be nested within eachother. These `universalmotif_df`s
form the basis for a majority of `memes` data outputs. These structures are also
valid input types to `memes` functions, and when used as such, output data are
appended as new columns to the input data, ensuring data provenance. Finally,
because `universalmotif_df`s are extensions of base R data.frames, they can be
manipulated using base R and `tidyverse` workflows.  Therefore, `memes` data
integrate seamlessly with common R workflows.
